Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-23 write-ups/FSE/paper.tex
Small changes.

@@

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.

@@

@@

\myparagraph{Preemptive multitasking on GPUs}
%
Hardware support for preemption has been proposed for \nvidia GPUs,
as well as \emph{SM-draining}
whereby workgroups occupying a symmetric multiprocessor (SM; a compute
unit using our terminology) are allowed to complete until the SM becomes free for
other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited the presence of blocking constructs, since it may not be possible to drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing},
where a workgroup can be re-scheduled
from scratch if it has not yet committed side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
using simulators, over sets of regular GPU kernels.  Very recent \nvidia GPUs support
preemption, though it is not clear whether they guarantee fairness~\cite{PascalWhitepaper}.

>>>>

\ADComment{Update intro to be clearer that even with preemption our approach has merit, and that preemption does not imply fairness.}

\ADComment{Use NVIDIA/Pascal as motivation for our approach.}

@@

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.
\ADComment{Tyler: briefly state that this still occurs on Pascal, despite it having support for preemption.}

@@

\subsection{Comparison with Preemption on NVIDIA GPUs}\label{sec:nvidiacomparison}

\ADComment{Make the point that NVIDIA GPUs now support
  instruction-level preemption in hardware; allows arbitrary
  preemption.}

\ADComment{Note that they are doing coarse-grained, task-level
  preemption, rather than running multiple tasks concurrently on the
  GPU.}

\ADComment{Direct comparison is not possible as we cannot implement
  cooperative kernels on NVIDIA devices due to lack of mature OpenCL
  2.0 support.}

\ADComment{However, based on results reported by NVIDIA and a clock
  speed calculation we can estimate...}

@@

\myparagraph{Preemptive multitasking on GPUs}
%
\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
%
Hardware support for preemption has been proposed for \nvidia GPUs,
as well as \emph{SM-draining}
whereby workgroups occupying a symmetric multiprocessor (SM; a compute
unit using our terminology) are allowed to complete until the SM becomes free for
other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited the presence of blocking constructs, since it may not be possible to drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing},
where a workgroup can be re-scheduled
from scratch if it has not yet committed side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
using simulators, over sets of regular GPU kernels.  Very recent \nvidia GPUs support
preemption, though it is not clear whether they guarantee fairness~\cite{PascalWhitepaper}.
